### 2014-06-06 writing/S'14 paper/join.tex
Typo
Proofreading

\begin{figure}[htp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \newcommand{\labelednode}[4]{\put(#1,#2){\oval(1.5,1)}\put(#1,#2){\makebox(0,0){$\begin{array}{c}#3\\\{#4\}\end{array}$}}}
    \setlength{\unitlength}{1cm}
    \begin{picture}(5,5.5)
      \labelednode{2.50}{5}{}{1,2,3}

@@

In our experiments, we create 80 randomly generated sets drawing from
a domain of seven elements. This yields a data set consisting of
6400 statements about pairs of entities. 3200 of these pairs are
chosen as a test set, and that test set is further reduced to the 2960
examples that can be provably derived from the training data.

@@

These results are fairly straightforward to interpret. The RNTN model
was able to accurately encode the relations between the terms in the
geometric relations between their vectors, and was able to then use
that information to recover relations that were not overtly included
in the training data. In contrast, the RNN model was able to
approximate this behavior only incompletely. It is possible but not
likely that it could be made to find a good solution with further
optimization on different learning algorithms, or that it would do
better on a larger universe of sets for which there would be a larger
set of training data to learn from, but the RNTN is readily able to achieve
these effects in the setting discussed here.

>>>>

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \newcommand{\labelednode}[4]{\put(#1,#2){\oval(1.5,1)}\put(#1,#2){\makebox(0,0){$\begin{array}{c}#3\\\{#4\}\end{array}$}}}
    \setlength{\unitlength}{1cm}
    \begin{picture}(5,5.5)
      \labelednode{2.50}{5}{}{1,2,3}

@@

In our experiments, we create 80 randomly generated sets drawing from
a domain of seven elements. This yields a data set consisting of
6400 statements about pairs of formulae. 3200 of these pairs are
chosen as a test set, and that test set is further reduced to the 2960
examples that can be provably derived from the training data.

@@

These results are fairly straightforward to interpret. The RNTN model
was able to accurately encode the relations between the terms in the
geometric relations between their vectors, and was able to then use
that information to recover relations that were not overtly included
in the training data. In contrast, the RNN model was able to achieve
this behavior only incompletely. It is possible but not likely that it
could be made to find a good solution with further optimization on
different learning algorithms, or that it would do better on a larger
universe of sets, for which there would be a larger set of training
data to learn from, but the RNTN is readily able to achieve these
effects in the setting discussed here.

### 2014-06-06 writing/S'14 paper/methods.tex
Proofreading

We study neural models that adhere to the linguistic \ii{principle of
 compositionality}, which says that the meanings for complex
expressions are derived from the meanings of their constituent parts
via specific combination functions \cite{Partee84,Janssen97}. In our
distributed setting, word meanings are vectors of length $N$. The
combination function maps pairs of them to single vectors of length $N$
vectors, which can then be merged again to create more complex
phrases. Once the entire sentence-level representation has been
derived, it is fed to a classifier trained in a supervised manner.

@@

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the RNTN layer function
\eqref{rntn} proposed in \citet{chen2013learning}. A sigmoid
nonlinearity (element-wise $\tanh$) is applied to the output of either
layer function, following \cite{socher2013acl1}.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}) \\ % TODO: Add column vectors?
\label{rntn}
\vec{y}_{\textit{RNTN}} = f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)} + \mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b})
\end{gather} 
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $N \times 2N$ matrix of learned weights, and applies the
element-wise non-linearity to the resulting vector. The RNTN has the
same basic structure, but with the addition of a learned third-order
tensor $\mathbf{T}$, dimension $N \times N \times N$, modeling
multiplicative interactions between the child vectors. Both models
include a bias vector, represented as $\vec{b}$.

>>>>

We study neural models that adhere to the linguistic \ii{principle of
 compositionality}, which says that the meanings for complex
expressions are derived from the meanings of their constituent parts
via specific combination functions \cite{Partee84,Janssen97}. In our
distributed setting, word meanings are vectors of length $n$. The
combination function concatenates pairs of them to form vectors of
length $2n$ and maps those larger vectors back into length $n$
vectors, which can then be merged again to create more complex
phrases. Once the entire sentence-level representation has been
derived, it is fed to a classifier trained in a supervised manner.

@@

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the RNTN layer function
\eqref{rntn} proposed in \citet{chen2013learning}. A sigmoid
nonlinearity (element-wise $\tanh$) is applied to the output of either
layer function, following \cite{socher2013acl1}.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}) \\ % TODO: Add column vectors?
\label{rntn}
\vec{y}_{\textit{RNTN}} = f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)} + \mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b})
\end{gather} % TODO: Explain third order tensor parameter?
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $n \times 2n$ matrix of learned weights, and applies the
element-wise non-linearity to the resulting vector. The RNTN has the
same basic structure, but with the addition of a learned third-order
tensor $\mathbf{T}$, dimension $n \times n \times n$, modeling
multiplicative interactions between the child vectors. Both models
include a bias vector, represented as $\vec{b}$.

### 2014-06-06 writing/S'14 paper/quantifiers.tex
Proofreading

We have seen that the RNTN can learn an approximation of propositional
logic.  However, natural languages can express functional meanings of
considerably greater complexity than this.  As a first step towards
investigating whether our models can capture this complexity, we now
attempt to directly measure the degree to which RNNs are able to
develop suitable representations for the semantics of natural language
quantifiers like \ii{some} and \ii{all}. Quantification is far from
the only place in natural language where complex functional meanings
are found, but it is a natural starting point, since it can be tested
in sentences whose structures are otherwise quite simple, and since it
has formed a standard case study in prior formal work on natural
language inference.

@@

This experiment replicates similar work described in
\cite{bowman2013can}, which found that RNTNs can learn to reason well
with quantifier meanings given sufficient training data. This paper
replaces the partially manually annotated data in that paper with data
that is generated directly using the logical system that we hope to
model, yielding results that we believe to be substantially more
straightforward to interpret.

@@

We evaluate the model using two experimental settings. In the simpler
setting, \textsc{all split}, we randomly sample 85\% of the data and evaluate on the
remaining 15\%. In this setting, the model is being asked to learn a
complete reasoning system for the limited language and logic presented
in the training data, but it is not being asked to generalize to test
examples that are substantially different from those it was trained
on. Crucially though, to succeed on this task, the model must be able
to recognize all of the lexical relations between the nouns, verbs,
and quantifiers and how they interact. For instance, it might see
\eqref{p1} and \eqref{p2} in training and, from that information,
determine \eqref{p3}.
%
% do not allow a blank line --- adds too much space
%
\begin{gather}
  \text{(most turtle) swim} \natalt \text{(no turtle) move}\label{p1}
  \\
  \text{(all lizard) reptile} \natfor  \text{(some lizard) animal}\label{p2}
  \\
  \text{(most turtle) reptile} \natalt \text{(no turtle) animal}\label{p3}
\end{gather}
%
% do not allow a blank line --- adds too much space
%
While our primary interest is in discovering the extent to which our
models can learn to encode the logic given an arbitrary amount of
data, we are also interested in the degree to which they can infer a
correct representation for the logic from more constrained training
data. To this end, we segment the sentence pairs according to which
quantifiers appear in each pair, and then hold out one such pair for
testing. We hypothesize that a model that can efficiently learn to
represent a logic should be able to construct an accurate
representation of each held out quantifier from the way that it
interacts with the other nine quantifiers which are not held
out. Since running this experiment requires choosing a pair of
quantifiers to hold out before training, the resource demands of
training prevent us from testing each of the 55 possible possible
pairs of quantifiers, and we choose only four pairs to test on.  Three
of these (\ii{two}/\ii{less-than-two}, \ii{not-all}/\ii{not-most}, and
\ii{all}/\ii{some}) were chosen because they allow for the most
different class labels at in the training data. The fourth is a
self-pair (\ii{no}/\ii{no}), meant to test that the model correctly
handles equality.

\begin{table}[tp]
  \centering
  \begin{tabular}{ l rrr }
    \toprule
    Data & Most frequent class & 20 dim RNN (\%) & 20 dim RNTN (\%)\\
    \midrule
    \textsc{all split}	& 35.4 &	65.3&	\textbf{99.0}
    \\[1ex]    
    \textsc{pair two/less-than-two}	& 59.8 &	77.6 &	\textbf{100.0} \\
    \textsc{pair not-all/not-most}	&0 & 60.8 &	\textbf{82.0} \\
    \textsc{pair all/some}	& 0& 52.5  &	\textbf{85.1} \\
    \textsc{pair no/no}	& 30.8 &	68.9 &	\textbf{99.9} \\
    \bottomrule
  \end{tabular}
  \caption{Performance on the quantifier experiments. }
  \label{resultstable}
\end{table}

\paragraph{Results} 
The results for these experiments are shown in
Figure~\ref{resultstable}. We compare the results to a most frequent
class baseline, which reflects the frequency in the test data of most
frequent class in the training data, $\natind$.  After some
cross-validation, we chose 20-dimensional word and phrase
representations, and 75 dimensional feature vectors for the
classifier.

The RNN performed poorly at this task, even though the sentences used
in these examples are short enough to avoid the pathology shown in
Figure~\ref{prop-falloff}.  However, the perfect performance by the
RNTN on the \textsc{pair no/no} experiment and its strong performance
on \textsc{all split} and \textsc{pair two/less-than-two} suggests
that that this stronger model is able to handle quantifiers correctly
given sufficient training data, but the sub-90\% results on two of the
training settings suggest that it may not be able to generalize from
data this impoverished in general. However, though the question of how
much data is necessary to accurately capture quantifier behavior in a
na\"ive model remains open, the fact that both models perform far
above baseline is promising.

>>>>

We have seen that the RNTN can learn an approximation of propositional
logic.  However, natural languages can express functional meanings of
considerably greater complexity than this.  As a first step towards
investigating whether our models can capture this complexity, we now
attempt to directly measure the degree to which they are able to
develop suitable representations for the semantics of natural language
quantifiers like \ii{some} and \ii{all}. Quantification is far from
the only place in natural language where complex functional meanings
are found, but it is a natural starting point, since it can be tested
in sentences whose structures are otherwise quite simple, and since it
has formed a standard case study in prior formal work on natural
language inference \cite{Icard:Moss:2013:LILT}.

@@

This experiment replicates similar work described in
\cite{bowman2013can}, which found that RNTNs can learn to reason well
with quantifier meanings given sufficient training data. This paper
replaces the partially manually annotated data in that paper with data
that is generated directly using the logical system that we hope to
learn, yielding results that we believe to be substantially more
straightforward to interpret.

@@

We evaluate the model using two experimental settings. In the simpler
setting, \textsc{all split}, we randomly sample 85\% of the data and evaluate on the
remaining 15\%. In this setting, the model is being asked to learn a
complete reasoning system for the limited language and logic presented
in the training data, but it is not being asked to generalize to test
examples that are substantially different from those it was trained
on. Crucially, though, to succeed on this task, the model must be able
to recognize all of the lexical relations between the nouns, verbs,
and quantifiers and how they interact. For instance, it might see
\eqref{p1} and \eqref{p2} in training and, from that information,
determine \eqref{p3}.
%
% do not allow a blank line --- adds too much space
%
\begin{gather}
  \text{(most turtle) swim} \natalt \text{(no turtle) move}\label{p1}
  \\
  \text{(all lizard) reptile} \natfor  \text{(some lizard) animal}\label{p2}
  \\
  \text{(most turtle) reptile} \natalt \text{(no turtle) animal}\label{p3}
\end{gather}
%
% do not allow a blank line --- adds too much space
%
While our primary interest is in discovering the extent to which our
models can learn to encode the logic given an arbitrary amount of
data, we are also interested in the degree to which they can infer a
correct representation for the logic from more constrained training
data. To this end, we segment the sentence pairs according to which
quantifiers appear in each pair, and then hold out one such pair for
testing. We hypothesize that a model that can efficiently learn to
represent a logic should be able to construct an accurate
representation of each held out quantifier from the way that it
interacts with the other nine quantifiers which are not held
out. Since running this experiment requires choosing a pair of
quantifiers to hold out before training, the resource demands of
training prevent us from testing each of the 55 possible 
pairs of quantifiers, so we choose only four pairs to test on.  Three
of these (\ii{two}/\ii{less-than-two}, \ii{not-all}/\ii{not-most}, and
\ii{all}/\ii{some}) were chosen because they allow for the most
different class labels in the training data. The fourth is a
self-pair (\ii{no}/\ii{no}), meant to test that the model correctly
handles equality.

\begin{table}[tp]
  \centering
  \setlength{\tabcolsep}{10pt}
  \begin{tabular}{ l r r@{ \ }r r@{ \ }r }
    \toprule
    Data & Most frequent class & \multicolumn{2}{c}{16 dim RNN}  & \multicolumn{2}{c}{20 dim RNTN}\\
    \midrule
    \textsc{all split}	            & 35.4 (7.5) &	67.4&(56.5)& \textbf{100.0} & \textbf{(100.0)}
    \\[1ex]    
    \textsc{pair two/less-than-two} & 59.8 (18.4) & 77.2 &(57.4) &	\textbf{100.0} &\textbf{(100.0)} \\
    \textsc{pair not-all/not-most}  &    0 (0)    & 66.0 &(54.3) &	\textbf{93.8}  &\textbf{(91.5)} \\
    \textsc{pair all/some}	    &    0 (0)    & 62.7 &(65.9)  &	\textbf{78.4}  &\textbf{(84.7)} \\
    \textsc{pair no/no}	            & 30.8  (10.2) & 67.8 & (61.8) &	\textbf{99.9}  &\textbf{(99.7)} \\
    \bottomrule
  \end{tabular}
  \caption{Performance on the quantifier experiments. Results are reported as accuracy scores followed by macroaveraged F1 scores in parentheses.}
  \label{resultstable}
\end{table}

\paragraph{Results} 
The results for these experiments are shown in
Figure~\ref{resultstable}. We compare the results to a most frequent
class baseline, which reflects the frequency in the test data of the
most frequent class in the training data, $\natind$.  After some
cross-validation, we chose 16- and 20-dimensional representations for
the RNN and RNTN, respectively, and 75-dimensional feature vectors for
the classifier.

The RNN performed poorly at this task, even though the sentences used
in these examples are short enough to avoid the pathology shown in
Figure~\ref{prop-falloff}.  However, the perfect performance by the
RNTN on the \textsc{all split} and \textsc{pair two/less-than-two}
experiments, and its strong performance on the \textsc{pair no/no}
experiment, suggest that this stronger model is able to handle
quantifiers correctly given sufficient training data, but the weaker
results on two of the training settings suggest that it may not be
able to generalize from data in general. However, though the question
of how much data is necessary to accurately capture quantifier
behavior in a na\"ive model remains open, the fact that both models
perform far above the baseline is promising.

### 2014-06-06 writing/S'14 paper/recursion.tex
Proofreading

\paragraph{Experiments}
As in Section~\ref{sec:join}, we test this phenomenon within the
framework of natural logic, but we now replace the unanalyzed symbols
from that experiment with complex formulae. These formulae
represent a truth-functionally complete classical propositional logic:
each atomic variable denotes either $\True$ or $\False$, and the only
operators are truth-functional ones.  Table~\ref{tab:pl} defines this
logic, and Table~\ref{tab:plexs} gives some examples of relational
statements that we can make in these terms. To compute these relations
between statements, we exhaustively enumerate the sets of assignments
of truth values to proposition variables that would satisfy each of
the statements and then convert the set-theoretic relation between
those assignments into one of the seven relations in
Table~\ref{b-table}. As a result, each relational statement represents
a valid theorem of the propositional logic.

@@

% NOTE: There was some confusion at CSLI and the 224U lecture about
% whether this was theorem proving or some kind of model checking.
% Hopefully this is explicit enough.
Socher et al.~\cite{socher2012semantic} show that a matrix-vector RNN
model somewhat similar to our RNTN can learn a logic, but the logic
discussed here is a much better approximation of the kind of
expressivity needed for natural language. The logic learned in that
experiment is boolean, wherein the atomic symbols are simply the
values $0$ and $1$, rather than variables over those values. While
learning the operators of that logic is not trivial, the outputs of
each operator can be represented accurately by a single bit. The
statements of propositional logic learned here describe conditions on
unknown truth values of propositions. As opposed to the two-way
contrasts seen in \cite{socher2012semantic}, this logic distinguishes
between $2^{6} = 64$ possible assignments of truth values, and
expressions of this logic define arbitrary conditions on these
possible assignments, for a total of $2^{64}$ %($\approx 10^{20}$)
possible statements that the intermediate vector representations need
to be able to distinguish if it is to learn this logical system in its
full generality.

@@

If we do not implement any constraint that the two statements being
compared are similar in any way, then the generated data are dominated
by statements in which the two formulae refer to largely separate
subsets of the six variables, which means that the $\natind$ relation
is almost always correct.  In an effort to balance the distribution of
relation labels without departing from the basic task of modeling
propositional logic, we disallow individual pairs of statements from
referring to more than four of the six proposition variables.

\paragraph{Results} 
We trained both the RNN and RNTN models on data with up to 4
connectives (65k pairs) and tested it on examples of up to 12
connectives (44k pairs). We initialized the model parameters randomly,
including the vector representations of the six variables. The results
are shown in Figure~\ref{prop-results}. In tuning, we found that the
RNN model was approximately optimal with 45-dimensional vector
representations, and the RNTN model was approximately optimal with 25
dimensions. We fixed the size of the feature vector for the classifier
at 75 dimensions. We found that the RNTN model was able to perform
almost perfectly on unseen small test examples, with accuracy above
98.7\% below size four (macro-F1 was above 95.5).  Starting at size
four, performance gradually falls with increasing size.  The RNN model
did not perform well, reaching only 90.5\% (81.4 micro-F1) accuracy on
the smallest test examples, and declining from there to near-baseline
performance at size 12. Training accuracy was 99.6\% for the RNTN and
87.6\% for the RNN.

>>>>

\paragraph{Experiments}
As in Section~\ref{sec:join}, we test this phenomenon within the
framework of natural logic, but we now replace the unanalyzed symbols
from that experiment with complex formulae. These formulae
represent a truth-functionally complete classical propositional logic:
each atomic variable denotes either $\True$ or $\False$, and the only
operators are truth-functional ones.  Table~\ref{tab:pl} defines this
logic, and Table~\ref{tab:plexs} gives some examples of relational
statements that we can make in these terms. To compute these relations
between statements, we exhaustively enumerate the sets of assignments
of truth values to propositional variables that would satisfy each of
the statements, and then we convert the set-theoretic relation between
those assignments into one of the seven relations in
Table~\ref{b-table}. As a result, each relational statement represents
a valid theorem of the propositional logic.

@@

% NOTE: There was some confusion at CSLI and the 224U lecture about
% whether this was theorem proving or some kind of model checking.
% Hopefully this is explicit enough.
Socher et al.~\cite{socher2012semantic} show that a matrix-vector RNN
model somewhat similar to our RNTN can learn a logic, but the logic
discussed here is a much better approximation of the kind of
expressivity needed for natural language. The logic learned in that
experiment is boolean, wherein the atomic symbols are simply the
values $0$ and $1$, rather than variables over those values. While
learning the operators of that logic is not trivial, the outputs of
each operator can be represented accurately by a single bit. The
statements of propositional logic learned here describe conditions on
unknown truth values of propositions. As opposed to the two-way
contrasts seen in \cite{socher2012semantic}, this logic distinguishes
between $2^{6} = 64$ possible assignments of truth values, and
expressions of this logic define arbitrary conditions on these
possible assignments, for a total of $2^{64}$ %($\approx 10^{20}$)
possible statements that the intermediate vector representations need
to be able to distinguish if we are to learn this logical system in its
full generality.

@@

If we do not implement any constraint that the two statements being
compared are similar in any way, then the generated data are dominated
by statements in which the two formulae refer to largely separate
subsets of the six variables, which means that the $\natind$ relation
is almost always correct.  In an effort to balance the distribution of
relation labels without departing from the basic task of modeling
propositional logic, we disallow individual pairs of statements from
referring to more than four of the six propositional variables.

\paragraph{Results} 
We trained both the RNN and RNTN models on data with up to 4
connectives (65k pairs) and tested it on examples of up to 12
connectives (44k pairs). We initialized the model parameters randomly,
including the vector representations of the six variables. The results
are shown in Figure~\ref{prop-results}. In tuning, we found that the
RNN model was approximately optimal with 45-dimensional vector
representations, and the RNTN model was approximately optimal with 25
dimensions. We fixed the size of the feature vector for the classifier
at 75 dimensions. We found that the RNTN model was able to perform
almost perfectly on unseen small test examples, with accuracy above
98.7\% on formulae below length four (macro-F1 was above 95.5).
Starting at size four, performance gradually falls with increasing
size.  The RNN model did not perform well, reaching only 90.5\% (81.4
micro-F1) accuracy on the smallest test examples, and declining from
there to near-baseline performance on formulae of length 12. Training
accuracy was 99.6\% for the RNTN and 87.6\% for the RNN.
